---
title: "BUS 4440_CDC"
author: "C. Krance"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include=TRUE, 
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      comment = " ", 
                      eval = TRUE)
```

Coronary heart disease (CHD) is the leading cause of death in the United States, 
accounting for approximately 375,000 deaths annually. Despite its prevalence, CHD 
remains significantly underdiagnosed, particularly among populations with limited 
healthcare access, creating opportunities for data-driven screening tools that can 
identify high-risk individuals before a clinical event occurs. Early and accurate 
classification of CHD risk has direct applications in preventive care, patient 
stratification for clinical trial enrollment, and targeted pharmaceutical intervention.

This project uses data from the 2024 Behavioral Risk Factor Surveillance System 
(BRFSS), an annual CDC telephone survey collecting health-related behavioral and 
chronic condition data from over 400,000 U.S. adults. The BRFSS is one of the 
largest continuously conducted health surveys in the world and serves as a primary 
data source for population-level disease surveillance. Variables were selected to 
capture three domains known to be associated with CHD risk: lifestyle behaviors 
(smoking, physical activity, alcohol consumption), chronic condition history 
(diabetes, stroke), and demographic characteristics (age, sex, race, BMI, income, 
education).

Three classification models (decision tree, logistic regression, and XGBoost) were 
trained and evaluated on a downsampled, balanced dataset to address the natural 
class imbalance in CHD prevalence. LASSO regularization was applied for variable 
selection prior to model training. Models were evaluated on a held-out validation 
set using sensitivity, specificity, and AUC as primary performance metrics, with 
sensitivity prioritized given the clinical cost asymmetry between false negatives 
and false positives in a medical screening context.


### Hypothesis
This project examines whether demographic, lifestyle, and chronic condition variables from the 2024 BRFSS can accurately classify coronary heart disease status. I hypothesize that logistic regression and ensemble models trained on these variables will achieve AUC above 0.80, with age, stroke history, and self-reported general health emerging as the strongest predictors.



```{r}
library(haven)
library(tidyverse)
library(dplyr)
library(caret)
library(rpart)
library(rpart.plot)
library(glmnet) #lasso
library(ggplot2)
library(pROC)
library(xgboost)
library(Ckmeans.1d.dp)
library(doParallel)

```

```{r}
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)
```


```{r}
df_raw <- read_xpt("LLCP2024.XPT")%>% 
  dplyr::select(
    `_MICHD`,     # outcome — reported having CHD or MI
    `_SEX`,       # biological sex
    `_AGEG5YR`,   # age categories
    `_RACE`,      # race categories
    `_BMI5CAT`,   # BMI categories
    `_EDUCAG`,    # education categories
    `_INCOMG1`,   # income brackets
    SMOKE100,     # smoked 100 cigs in lifetime
    CVDSTRK3,     # ever had a stroke
    DIABETE4,     # diabetes (Yes/No)
    GENHLTH,      # self assessed general health categories
    MEDCOST1,     # can't see doctor due to cost
    CHECKUP1,     # time since last routine checkup
    `_TOTINDA`,   # physical activity last 30 days
    AVEDRNK4,     # avg drinks over last 30 days
    `_HLTHPL2`,   # health insurance coverage
    MENTHLTH      # mental health bad days
  )

```



```{r}
df_clean <- df_raw %>% 
  mutate(
    MICHD = case_when(
                        `_MICHD` == 2 ~ 0,
                        `_MICHD` == 1 ~ 1,
                        TRUE ~ NA_real_
    ),
      SEX = factor(`_SEX`,
                      levels = 1:2,
                      labels = c("Male", "Female")
       
    ),
      AGEG5YR = factor(`_AGEG5YR`,
                          levels = 1:13,
                          labels = c("18_24", "25_29", "30_34", "35_39", "40_44",
                                     "45_49", "50_54", "55_59", "60_64", "65_69",
                                     "70_74", "75_79", ">80")
    ),
      RACE = factor(`_RACE`,
                       levels = 1:8,
                       labels = c("White", "Black", "American Indian/Alaskan Native", 
                                  "Asian", "Native Hawaiian_Pacific Islander", 
                                  "Other", "Multiracial", "Hispanic")
    ),
      EDUCAG = factor(`_EDUCAG`,
                         levels = 1:4,
                         labels = c("DNF_High_School","High_School_Grad",
                                    "Attended_College","College_Grad")
                         
    ),
        BMI5CAT = factor(`_BMI5CAT`,
                            levels = 1:4,
                            labels = c("Underweight", "Normal", 
                                       "Overweight", "Obese")
    
    ),
      INCOMG1 = case_when(
                            `_INCOMG1` == 9 ~ NA_real_,
                            TRUE ~ `_INCOMG1`
    ),
      SMOKE100 = case_when(
                          SMOKE100 == 1 ~ 1,
                          SMOKE100 == 2 ~ 0,
                          TRUE ~ NA_real_
                      
    ),
      CVDSTRK3 = case_when(
                          CVDSTRK3 == 1 ~ 1,
                          CVDSTRK3 == 2 ~ 0,
                          TRUE ~ NA_real_
        
    ),
      DIABETE4 = case_when(
                          DIABETE4 == 1 ~ 1,
                          DIABETE4 %in% 2:4 ~ 0,
                          TRUE ~ NA_real_
        
    ),
      GENHLTH = factor(GENHLTH,
                      levels = 1:5,
                      labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")
                      
    ),
      MEDCOST1 = case_when(
                          MEDCOST1 == 1 ~ 1,
                          MEDCOST1 == 2 ~ 0,
                          TRUE ~ NA_real_
        
    ),
      CHECKUP1 = factor(CHECKUP1,
                        levels = c(1,2,3,4,8),
                        labels = c("<1_year","1_2_years","2_5_years",">5_years","Never")
        
    ),
      TOTINDA = case_when(
                            `_TOTINDA` == 1 ~ 1,
                            `_TOTINDA` == 2 ~ 0,
                            TRUE ~ NA_real_
        
    ),
      AVEDRNK4 = case_when(
                          AVEDRNK4 == 88 ~ 0,      ## coded as none
                          AVEDRNK4 %in% 1:76 ~ AVEDRNK4,
                          TRUE ~ NA_real_ 
                        
    ),
      HLTHPL2 = case_when(
                            `_HLTHPL2` == 1 ~ 1,
                            `_HLTHPL2` == 2 ~ 0,
                            TRUE~ NA_real_
        
    ),
      MENTHLTH = case_when(
                          MENTHLTH == 88 ~ 0,
                          MENTHLTH %in% 1:30 ~ MENTHLTH,
                          TRUE ~ NA_real_
      ),
  ) %>%
  dplyr::select(MICHD, everything(), -`_MICHD`, -`_SEX`, -`_AGEG5YR`, -`_RACE`, 
         -`_EDUCAG`, -`_BMI5CAT`, -`_INCOMG1`, 
         -`_TOTINDA`, -`_HLTHPL2`)
  
```



```{r}
df_clean <- df_clean %>%
  drop_na()
```



```{r}
set.seed(123)
df_clean_down <- downSample(
  x = df_clean %>% dplyr::select(-MICHD),     ## Addresses huge class imbalance
  y = as.factor(df_clean$MICHD),
  yname = "MICHD"
)
```


```{r}
set.seed(123)

idx <- createDataPartition(df_clean_down$MICHD, p = 0.6, list = FALSE)
train.df <- df_clean_down[idx,]
holdout.df <- df_clean_down[-idx,]
```



```{r}

# 1. Separate features and target
x_train <- train.df[, !names(train.df) %in% "MICHD"]
x_test  <- holdout.df[, !names(holdout.df) %in% "MICHD"]
y_train <- train.df$MICHD
y_test  <- holdout.df$MICHD

# 2. Dummy encode using model.matrix
x_train_matrix <- model.matrix(~ . - 1, data = x_train)
x_test_matrix  <- model.matrix(~ . - 1, data = x_test)

# 3. Fit Lasso
cv_lasso <- cv.glmnet(x_train_matrix, y_train, alpha = 1, family = "binomial", nfolds = 5)

# 4. Predict on holdout
predictions <- predict(cv_lasso, newx = x_test_matrix, s = "lambda.min", type = "class")
```



```{r, results='hide'} 


# 1. Confusion matrix
confusionMatrix(as.factor(predictions), as.factor(y_test))

# 2. Quick accuracy
mean(predictions == y_test)

# 3. See which variables Lasso kept (non-zero coefficients)
coef(cv_lasso, s = "lambda.min")

# 4. Plot CV error vs lambda
plot(cv_lasso)



```



```{r, results='hide'}
lasso_coefs <- coef(cv_lasso, s = "lambda.min")
lasso_df <- data.frame(
  variable = rownames(lasso_coefs),
  coefficient = as.numeric(lasso_coefs)
)

# Sort by absolute value to see most important variables
lasso_df <- lasso_df[order(abs(lasso_df$coefficient), decreasing = TRUE), ]
lasso_df <- lasso_df[lasso_df$coefficient != 0, ]  # drop zeroes
print(lasso_df)
```

Variables with the largest absolute effect were kept in the final dataset. 9 Explanatory variables performed nearly identically to all 16 variables.


```{r}
df_final <- df_clean_down %>% 
          dplyr::select(MICHD, AGEG5YR, CVDSTRK3, GENHLTH, SEX, CHECKUP1, RACE, DIABETE4, MEDCOST1, SMOKE100)
```


```{r}
set.seed(123)

idx <- createDataPartition(df_final$MICHD, p = 0.6, list = FALSE)
train.df <- df_final[idx,]
holdout.df <- df_final[-idx,]
```


```{r}
tree_model_1 <- rpart(MICHD ~ .,
                      data = train.df,
                      method = "class",
                      parms = list(prior = c(0.5, 0.5)),
                      control = rpart.control(
                        maxdepth = 10,    
                        minsplit = 20,    # min observations to attempt a split
                        cp = 0.001        # lower = more splits, more complex tree
                      ))
g = rpart.plot(tree_model_1, extra = 1, fallen.leaves = FALSE)
```


```{r, results='hide'}
png("decision_tree.png", width = 4000, height = 2500, res = 200)
rpart.plot(tree_model_1, 
           extra = 1, 
           fallen.leaves = FALSE,
           tweak = 1.2)
dev.off()
```



```{r, results='hide'}
tree.pred <- predict(tree_model_1, holdout.df, type = "class")
confusionMatrix(tree.pred, as.factor(holdout.df$MICHD), 
                positive = "1")
```
### Decision Tree

The tree visualizes that age and self-rated general health are the two strongest 
predictors of CHD, followed by stroke history and sex. The highest-risk profile 
is an older adult (55+, more prevalent in men) who rates their health as Fair or 
Poor and has a prior stroke history.

| Metric       | Value  |
|--------------|--------|
| Accuracy     | 75.66% |
| Sensitivity  | 81.34% |
| Specificity  | 69.98% |
| PPV          | 73.04% |
| NPV          | 78.95% |
| AUC          | 0.810  |

The model achieves the highest sensitivity of the three models, correctly identifying 
81.34% of true CHD cases. McNemar's test (p < 2.2e-16) confirms significantly more 
false positives than false negatives (1,385 vs. 861) — a clinically acceptable 
asymmetry, as over-flagging CHD risk is preferable to missing true cases. Performance 
is significantly better than random guessing (p < 2.2e-16).
  

```{r, results='hide'}
train.df$MICHD <- as.factor(train.df$MICHD)
holdout.df$MICHD <- as.factor(holdout.df$MICHD)

trControl <- caret::trainControl(method = "cv", number = 5, allowParallel = TRUE)
logit.reg <- caret::train(MICHD ~ ., data = train.df, trControl = trControl, 
                          method = "glm", family = "binomial")

logit.reg
summary(logit.reg$finalModel)
```



```{r, results='hide'}
logit.pred <- predict(logit.reg, holdout.df)
confusionMatrix(logit.pred, as.factor(holdout.df$MICHD), positive = "1")
```

### Logistic Regression

Age is the strongest predictor of CHD. Coefficients become positive and statistically 
significant beginning at ages 40–44 and increase monotonically through the 80+ group 
(β = 3.27, p < 0.001). Younger age groups were not statistically significant. 
Self-reported general health shows a strong dose-response pattern — Fair (β = 1.81) 
and Poor (β = 2.21) health carry the largest effects, though even "Very Good" and 
"Good" categories are positive and significant, suggesting CHD risk increases 
continuously as health declines.

Stroke history (β = 1.36, p < 0.001) is among the largest individual predictors, 
consistent with shared vascular pathology. Female sex (β = −0.82, p < 0.001) and 
Asian race (β = −0.72, p < 0.001) are negatively associated with CHD; most other 
race categories were not statistically significant. Time since last checkup shows an 
inverse relationship — individuals with checkups 2+ years ago have lower observed CHD 
odds (β = −0.70 for 5+ years, p < 0.001), though this likely reflects behavioral 
selection rather than a causal effect. Diabetes, cost barriers to care, and smoking 
are each positive and significant (p < 0.001).

| Metric       | Value  |
|--------------|--------|
| Accuracy     | 76.11% |
| Sensitivity  | 78.11% |
| Specificity  | 74.10% |
| PPV          | 75.10% |
| NPV          | 77.20% |
| AUC          | 0.839  |

Five-fold cross-validation produced a consistent accuracy of 76.42%, indicating the 
model generalizes well and is not overfit. McNemar's test (p < 0.001) confirms 
significantly more false positives than false negatives (1,195 vs. 1,010) — a 
clinically acceptable asymmetry for a screening application. Performance is 
significantly better than random guessing (p < 2.2e-16).



### XGBoost

```{r, results='hide'}

# Encode for XGBoost
x_train_xgb <- model.matrix(MICHD ~ . - 1, data = train.df)
x_hold_xgb  <- model.matrix(MICHD ~ . - 1, data = holdout.df)
y_train_xgb <- as.numeric(as.character(train.df$MICHD))
y_hold_xgb  <- as.numeric(as.character(holdout.df$MICHD))

dtrain <- xgb.DMatrix(data = x_train_xgb, label = y_train_xgb)
dhold  <- xgb.DMatrix(data = x_hold_xgb,  label = y_hold_xgb)

# Train — early stopping prevents overfitting and saves time
set.seed(123)
xgb_model <- xgb.train(
  params = list(
    objective        = "binary:logistic",
    eval_metric      = "auc",
    max_depth        = 6,
    eta              = 0.05,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    nthread          = parallel::detectCores() - 1
  ),
  data              = dtrain,
  nrounds           = 500,
  watchlist         = list(train = dtrain, eval = dhold),
  early_stopping_rounds = 20,   # stops if AUC doesn't improve for 20 rounds
  verbose           = 0
)

# Predict
xgb_prob  <- predict(xgb_model, dhold)
xgb_class <- as.factor(ifelse(xgb_prob > 0.5, 1, 0))

confusionMatrix(xgb_class, as.factor(holdout.df$MICHD), positive = "1")
```

```{r}
# Variable importance plot
xgb_imp <- xgb.importance(model = xgb_model)
xgb.ggplot.importance(xgb_imp, top_n = 15) +
  labs(title = "XGBoost Feature Importance") +
  theme_minimal(base_size = 12)
```

The XGBoost feature importance plot confirms the patterns identified by logistic 
regression and LASSO: the 80+ age group is the single most important feature, 
followed by diabetes, stroke history, and older age brackets. Female sex, declining 
general health, and smoking round out the top predictors. The close alignment between 
XGBoost and logistic regression variable importance provides independent validation 
that these relationships are robust and not model-specific artifacts.

| Metric       | Value  |
|--------------|--------|
| Accuracy     | 75.88% |
| Sensitivity  | 78.93% |
| Specificity  | 72.82% |
| PPV          | 74.39% |
| NPV          | 77.56% |
| AUC          | 0.838  |

The model was trained with early stopping (20 rounds) and regularization via 
subsampling and column sampling to prevent overfitting. McNemar's test 
(p < 2.587e-09) confirms significantly more false positives than false negatives 
(1,254 vs. 972) — consistent with the other models and clinically acceptable for 
a screening context. The marginal performance gap between XGBoost and logistic 
regression (ΔAUC = 0.001) suggests the underlying predictor relationships in this 
dataset are largely linear and well-captured by the simpler model. Performance is 
significantly better than random guessing (p < 2.2e-16).



```{r}
# Decision tree
tree_prob <- predict(tree_model_1, holdout.df, type = "prob")[, "1"]

# Logistic regression
logit_prob <- predict(logit.reg, holdout.df, type = "prob")[, "1"]


# Build ROC objects
roc_tree  <- roc(as.numeric(as.character(holdout.df$MICHD)), tree_prob,  quiet = TRUE)
roc_logit <- roc(as.numeric(as.character(holdout.df$MICHD)), logit_prob, quiet = TRUE)
roc_xgb   <- roc(as.numeric(as.character(holdout.df$MICHD)), xgb_prob,   quiet = TRUE)
```

```{r,`include=TRUE`}
# Agregated Chart
ggroc(list(
  "Decision Tree"       = roc_tree,
  "Logistic Regression" = roc_logit,
  "XGBoost"             = roc_xgb
), linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("#E63946", "#457B9D", "#2A9D8F")) +
  labs(
    title = "ROC Curves — CHD Classification Models",
    x = "Specificity", y = "Sensitivity", color = "Model"
  ) +
 annotate("text", x = 0.75, y = 0.05,
         label = paste0("AUC:  Tree: ",     sprintf("%.3f", auc(roc_tree)),
                        "   |   Logit: ",   sprintf("%.3f", auc(roc_logit)),
                        "   |   XGBoost: ", sprintf("%.3f", auc(roc_xgb))),
         size = 3.2, hjust = 0)
```


```{r}
model_comparison <- data.frame(
  Model          = c("Decision Tree", "Logistic Regression", "XGBoost"),
  Accuracy       = c(0.7566, 0.7611, 0.7588),
  Sensitivity    = c(0.8134, 0.7811, 0.7893),
  Specificity    = c(0.6998, 0.7410, 0.7282),
  Pos_Pred_Value = c(0.7304, 0.7510, 0.7439),
  Neg_Pred_Value = c(0.7895, 0.7720, 0.7756),
  Kappa          = c(0.5132, 0.5221, 0.5176),
  AUC            = c(0.810,  0.839,  0.838)
)
knitr::kable(model_comparison, caption = "Model Performance Comparison")
```




The ROC curves confirm that logistic regression is the best-performing model, achieving the 
highest AUC of 0.839, followed closely by XGBoost (0.838) and the decision tree (0.810). 
AUC evaluates model performance across all possible classification thresholds simultaneously, 
making it a more reliable basis for model comparison than single-threshold metrics like 
sensitivity or accuracy alone.

While the decision tree achieved the highest sensitivity (81.34%) at the default 0.50 
threshold, the ROC curve reveals this was partly an artifact of that threshold rather than 
superior overall discrimination. At any fixed operating point on the curve, logistic 
regression outperforms the decision tree while generating fewer false alarms.

Logistic regression was selected as the final model. It achieves the highest AUC, produces 
fully interpretable coefficients that map directly to known clinical risk factors, and 
matches XGBoost performance without the added complexity — confirming that on structured 
survey data of this type, a well-specified logistic regression is competitive with ensemble 
methods. All three models substantially outperform the no-information baseline (AUC = 0.50), 
and results are consistent with published CHD risk models such as the Framingham Heart Score, 
which achieves AUC in the 0.75–0.85 range on comparable population-level data.



```{r}
# Refit final model on full downsampled dataset for deployment
set.seed(123)
final_logit <- caret::train(MICHD ~ ., 
                            data = df_final, 
                            method = "glm", 
                            family = "binomial",
                            trControl = trainControl(method = "cv", 
                                                     number = 5, 
                                                     allowParallel = TRUE))
```

```{r}
# End parallel processing
stopCluster(cl)
```

The final logistic regression model was refit on the full downsampled dataset for deployment; 
holdout performance metrics reported above remain the authoritative evaluation.


### Limitations

Several limitations should be considered when interpreting these results. 
First, the BRFSS relies entirely on self-reported data, which introduces 
recall bias and potential misclassification. Respondents may inaccurately 
report diagnoses, behaviors, or health status. Second, the survey is 
cross-sectional, meaning all variables are measured at a single point in 
time. This prevents any causal inference; observed associations reflect 
correlation rather than directionality or mechanism.

Third, and most importantly for clinical applicability, the models are 
constrained to variables available through public health surveys. 
Established clinical biomarkers with strong predictive value for CHD, 
including LDL and HDL cholesterol, C-reactive protein, troponin levels, 
coronary artery calcium scores, and blood pressure readings, are absent 
from the BRFSS by design. The inclusion of even a small number of these 
biomarkers would likely improve AUC substantially and bring the model 
closer to the performance of validated clinical risk tools such as the 
Pooled Cohort Equations or HEART score. The current models should 
therefore be interpreted as population-level screening tools rather than 
clinical diagnostic instruments.

Finally, downsampling was used to address class imbalance, which 
improves model balance but reduces the effective training sample and may 
not reflect the true population prevalence of CHD.


### Acknowledgements

I would like to personally thank Dr. Shankar Ghimire at Florida Southern College for 
providing this opportunity as part of the BUS 4440 Data Mining course.









